---
title: "Lab04_DeepLearning_iNaturalist"
author: "Julia Parish"
date: "2022/03/02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Deep Learning - iNaturalist

Homework task is to apply deep learning skills to build the following models:

- 2 Species (binary classification) - neural net. Draw from 3.4 🍿 Movies (binary classification). You’ll need to pre-process the images to be a consistent shape first though – see 5.2.4 Data preprocessing.
- 2 Species (binary classification) - convolutional neural net. Draw from the dogs vs cats example.
- 10 Species (multi-class classification) - neural net. Draw from 3.5 📰 Newswires (multi-class classification).
- 10 Species (multi-class classification) - convolutional neural net. Draw from dogs vs cats example and update necessary values to go from binary to mult-class classification.

In the models, be sure to include the following:

- Split the original images per species (n=50) into train (n=30), validate (n=10) and test (n=10). These are almost absurdly few files to feed into these complex deep learning models but will serve as a good learning example.
- Include accuracy metric and validation in the fitting process and history plot.
- Evaluate loss and accuracy on your test model results. Compare standard neural network and convolutional neural network results.


```{r}
librarian::shelf(
  digest, dplyr, DT, keras, glue, here, readr, tensorflow, tidyverse)

# path to folder containing species directories of images
dir_src  <- "/courses/EDS232/inaturalist-2021/train_mini"
dir_dest <- here("inat/")
dir.create(dir_dest, showWarnings = F)

# get list of directories, one per species (n = 10,000 species)
dirs_spp <- list.dirs(dir_src, recursive = F, full.names = T)
n_spp <- length(dirs_spp)

# set seed (for reproducible results) 
# just before sampling (otherwise get different results)
# based on your username (unique amongst class)
Sys.info()[["user"]] %>% 
  digest::digest2int() %>% 
  set.seed()
i10 <- sample(1:n_spp, 10)

# show the 10 indices sampled of the 10,000 possible 
i10
```

```{r}
# show the 10 species directory names
basename(dirs_spp)[i10]
```

```{r}
# show the first 2 species directory names
i2 <- i10[1:2]
basename(dirs_spp)[i2]
```

### Split the original images per species (n=50) into train (n=30), validate (n=10) and test (n=10)

```{r}
# setup data frame with source (src) and destination (dest) paths to images
d <- tibble(
  set     = c(rep("spp2", 2), rep("spp10", 10)),
  dir_sp  = c(dirs_spp[i2], dirs_spp[i10]),
  tbl_img = map(dir_sp, function(dir_sp){
    tibble(
      src_img = list.files(dir_sp, full.names = T),
      subset  = c(rep("train", 30), rep("validation", 10), rep("test", 10))) })) %>% 
  unnest(tbl_img) %>% 
  mutate(
    sp       = basename(dir_sp),
    img      = basename(src_img),
    dest_img = glue("{dir_dest}/{set}/{subset}/{sp}/{img}"))

# show source and destination for first 10 rows of tibble
d %>% 
  select(src_img, dest_img)
```

```{r}
# iterate over rows, creating directory if needed and copying files 
d %>% 
  pwalk(function(src_img, dest_img, ...){
    dir.create(dirname(dest_img), recursive = T, showWarnings = F)
    file.copy(src_img, dest_img) })

# uncomment to show the entire tree of your destination directory
# system(glue("tree {dir_dest}"))
```

# 1. 2 Species (binary classification) - neural net

```{r}
train_dir_2spp <- here("inat/spp2/train")
test_dir_2spp <- here("inat/spp2/test")
validation_dir_2spp <- here("inat/spp2/validation")

# All images will be rescaled by 1/255
train_datagen_2spp <- image_data_generator(rescale = 1/255)
test_datagen_2spp <- image_data_generator(rescale = 1/255)
validation_datagen_2spp <- image_data_generator(rescale = 1/255)

train_generator_2spp <- flow_images_from_directory(
  train_dir_2spp,  # This is the target directory
  train_datagen_2spp, # This is the data generator
  target_size = c(150, 150), # All images will be resized to 150x150
  batch_size = 5,
  # Since we use binary_crossentropy loss, we need binary labels
  class_mode = "binary")

test_generator_2spp <- flow_images_from_directory(
  test_dir_2spp,
  test_datagen_2spp, 
  target_size = c(150, 150), 
  batch_size = 5,
  class_mode = "binary")

validation_generator_2spp <- flow_images_from_directory(
  validation_dir_2spp,
  validation_datagen_2spp,
  target_size = c(150, 150),
  batch_size = 5,
  class_mode = "binary")
```

```{r}
# The intermediate layers will use relu as their “activation function”, and the final layer will use a sigmoid activation so as to output a probability (a score between 0 and 1, indicating how likely the sample is to have the target “1”, i.e. how likely the review is to be positive). A relu (rectified linear unit) is a function meant to zero-out negative values, while a sigmoid “squashes” arbitrary values into the [0, 1] interval, thus outputting something that can be interpreted as a probability.

model_spp2 <- keras_model_sequential() %>% 
  layer_dense(units = 16, activation = "relu", input_shape = c(150, 150, 3)) %>% 
  layer_dense(units = 16, activation = "relu") %>% 
  layer_flatten() %>% 
  layer_dense(units =  1, activation = "sigmoid")
```

pick a loss function and an optimizer - 

```{r}
# we configure our model with the rmsprop optimizer and the binary_crossentropy loss function. Note that we will also monitor accuracy during training.

model_spp2 %>% compile(
  optimizer = "rmsprop",
  loss      = "binary_crossentropy",
  metrics   = c("accuracy"))
```

```{r}
# to configure the parameters of your optimizer 
model_spp2 %>% compile(
  optimizer = optimizer_rmsprop(learning_rate = 0.0001),
  loss      = "binary_crossentropy",
  metrics   = c("acc")) 

```

```{r}
# pass a custom loss function or metric function

# model_spp2 %>% compile(
#   optimizer = optimizer_rmsprop(lr = 0.001),
#   loss      = loss_binary_crossentropy,
#   metrics   = metric_binary_accuracy) 
```

```{r}
# train our model

history_spp2 <- model_spp2 %>% fit(
  train_generator_2spp,
  steps_per_epoch = 5,
  epochs = 10,
  validation_data = validation_generator_2spp,
  validation_steps = 1)

str(history_spp2)
```

```{r}
plot(history_spp2)
```

```{r}
results_spp2 <- model_spp2 %>% evaluate(test_generator_2spp)
results_spp2
```

## 2. 2 Species (binary classification) - convolutional neural net
 
```{r}
cnn_model_spp2 <- keras_model_sequential() %>% 
  layer_conv_2d(
    filters = 32, kernel_size = c(3, 3), activation = "relu",
    input_shape = c(150, 150, 3)) %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_conv_2d(filters = 64, kernel_size = c(3, 3), activation = "relu") %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_conv_2d(filters = 128, kernel_size = c(3, 3), activation = "relu") %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_conv_2d(filters = 128, kernel_size = c(3, 3), activation = "relu") %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_flatten() %>% 
  layer_dropout(rate = 0.5) %>% 
  layer_dense(units = 512, activation = "relu") %>% 
  layer_dense(units = 1, activation = "sigmoid")  
  
cnn_model_spp2 %>% compile(
  loss = "binary_crossentropy",
  optimizer = optimizer_rmsprop(learning_rate = 0.0001),
  metrics = c("acc"))

```

```{r}
summary(cnn_model_spp2)

```

#### Train Model

```{r}
# train our model

history_cnn_spp2 <- cnn_model_spp2 %>% fit(
  train_generator_2spp,
  steps_per_epoch = 5,
  epochs = 30,
  validation_data = validation_generator_2spp,
  validation_steps = 1)

str(history_cnn_spp2)
```

```{r}
plot(history_cnn_spp2)
```

```{r}
history_spp2
```


```{r}
history_cnn_spp2
```

```{r}
results_spp2cnn <- cnn_model_spp2 %>% evaluate(test_generator_2spp)
results_spp2cnn
```


# 3 10 Species (multi-class classification) - neural net. 


```{r}
train_dir_10spp <- here("inat/spp10/train")
test_dir_10spp <- here("inat/spp10/test")
validation_dir_10spp <- here("inat/spp10/validation")

# All images will be rescaled by 1/255
train_datagen_10spp <- image_data_generator(rescale = 1/255)
test_datagen_10spp <- image_data_generator(rescale = 1/255)
validation_datagen_10spp <- image_data_generator(rescale = 1/255)

train_generator_10spp <- flow_images_from_directory(
  train_dir_10spp,  # This is the target directory
  train_datagen_10spp, # This is the data generator
  target_size = c(150, 150), # All images will be resized to 150x150
  batch_size = 5,
  # 
  class_mode = "categorical")

test_generator_10spp <- flow_images_from_directory(
  test_dir_10spp,
  test_datagen_10spp,
  target_size = c(150, 150),
  batch_size = 5,
  class_mode = "categorical")

validation_generator_10spp <- flow_images_from_directory(
  validation_dir_10spp,
  validation_datagen_10spp,
  target_size = c(150, 150),
  batch_size = 5,
  class_mode = "categorical")
```


```{r}
model_10spp <- keras_model_sequential() %>% 
  layer_dense(units = 64, activation = "relu", input_shape = c(150, 150, 3)) %>% 
  layer_dense(units = 64, activation = "relu") %>% 
  layer_flatten() %>% 
  layer_dense(units =  10, activation = "softmax")

```

```{r}
# to configure the parameters of your optimizer 
model_10spp %>% compile(
  optimizer = optimizer_rmsprop(learning_rate = 0.0001),
  loss      = "categorical_crossentropy",
  metrics   = c("acc")) 

```

```{r}
# train our model

history_10spp <- model_10spp %>% fit(
  train_generator_10spp,
  steps_per_epoch = 5,
  epochs = 30,
  validation_data = validation_generator_10spp,
  validation_steps = 1)

```

```{r}
plot(history_10spp)
```

```{r}
results_10spp <- model_10spp %>% evaluate(test_generator_10spp)
results_10spp
```

```{r}
history_10spp
```

# 4. 10 Species (multi-class classification) - convolutional neural net

```{r}
cnn_model_10spp <- keras_model_sequential() %>% 
  layer_conv_2d(
    filters = 32, kernel_size = c(3, 3), activation = "relu",
    input_shape = c(150, 150, 3)) %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_conv_2d(filters = 64, kernel_size = c(3, 3), activation = "relu") %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_conv_2d(filters = 128, kernel_size = c(3, 3), activation = "relu") %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_conv_2d(filters = 128, kernel_size = c(3, 3), activation = "relu") %>% 
  layer_max_pooling_2d(pool_size = c(2, 2)) %>% 
  layer_flatten() %>% 
  layer_dropout(rate = 0.5) %>% 
  layer_dense(units = 512, activation = "relu") %>% 
  layer_dense(units = 10, activation = "softmax")  
  
```

```{r}
cnn_model_10spp %>% compile(
  loss = "categorical_crossentropy",
  optimizer = optimizer_rmsprop(learning_rate = 0.0001),
  metrics = c("acc"))
```

```{r}
history_cnn_10spp <- cnn_model_10spp %>% fit(
  train_generator_10spp,
  steps_per_epoch = 5,
  epochs = 30,
  validation_data = validation_generator_10spp,
  validation_steps = 10)
```

```{r}
history_cnn_10spp
```

```{r}
results_10sppcnn <- cnn_model_10spp %>% evaluate(test_generator_10spp)
results_10spp
```

## Standard neural net compared to convolutional neural net

```{r}
results_10spp
results_10sppcnn
```

The standard neural net model evaluated on the test images has a loss rate of `r results_10spp[[1]]` and an accuracy rate of `r results_10spp[[2]]`. The convolutional neural net model loss rate of `r results_10sppcnn[[1]]` and an accuracy rate of `r results_10sppcnn[[2]]`. These accuracy and loss rate show the standard neural network is more accurate. Typically, the convoluntional neural net is a better model, but not in this example as the data is too small. 


